The Impact of Alternation

### 2017-05-14 draft.tex
intro+related

>>>>

The language inclusion problem is recognized as being central to
verification of hardware, communication protocols and software
systems. A property is a specification of the correct executions of a
system, given as a set $\mathcal{P}$ of executions, and the
verification problem asks if the set $\mathcal{S}$ of executions of
the system under consideration is contained within $\mathcal{P}$.

This problem is at the core of widespread verification techniques,
such as automata-theoretic model checking \cite{VARDI94}, where
systems are specified as finite-state automata and properties defined
using Linear Temporal Logic \cite{Pnueli77}. However the bottleneck of
this and other related verification techniques is the intractability
of language inclusion (PSPACE-complete for finite-state automata on
finite words over finite alphabets).

Alternation \cite{ChandraKozenStockmeyer81} was introduced as a
generalization of nondeterminism, introducing universal, in addition
to existential transitions. For automata over finite alphabets, the
language inclusion problem can be encoded as the emptiness problem of
an alternating automaton of linear size. Moreover, efficient
exploration techniques based on antichains are shown to perform well
for alternating automata over finite alphabets \cite{DeWulf08}.

Using finite alphabets for the specification of properties and models
is however very restrictive, when dealing with real-life computer
systems, mostly because of the following reasons. On one hand,
programs handle data from very large domains, that can be assumed to
be infinite (64-bit integers, floating point numbers, strings of
characters, etc.) and their correctness must be specified in terms of
the data values. On the other hand, systems must respond to strict
deadlines, which requires temporal specifications as timed
languages \cite{AlurDill94}.

Although being convenient specification tools, automata over infinite
alphabets lack the decidability properties ensured by finite
alphabets. In general, when considering infinite data as part of the
input alphabet, language inclusion is undecidable and, even
complementation becomes impossible, for instance, for timed automata
\cite{AlurDill94} or finite-memory register automata
\cite{Kaminski94}. In some cases, one can recover theoretical
decidability, by restricting the number of variables (clocks) in timed
automata to one \cite{OuaknineWorrell04}, or forbidding relations
between current and past/future values, as with symbolic automata
\cite{symbTransd:POPL12}. In such cases, the emptiness problem for the
alternating versions becomes decidable \cite{Lasota05,DAntoniKW16}.

In this paper, we present a new model of alternating automata over
infinite alphabets consisting of pairs $(a,\nu)$ where $a$ is an input
event from a finite set and $\nu$ is a valuation of a finite set
$\vec{x}$ of variables that range over an infinite domain. We assume
that, at all times, the successive values taken by the variables in
$\vec{x}$ are an observable part of the language, in other words,
there are no hidden variables in our model. The transition rules are
specified by a set of formulae, in a combined first-order theory of
boolean control states and data, that relate past with present values
of the variables. We do not fix the data theory a~priori, but rather
consider it to be a parameter of the class.

A run over an input word $(a_1,\nu_1) \ldots (a_n,\nu_n)$ is a
sequence $\phi_0(\vec{x}_0) \Rightarrow \phi_1(\vec{x}_0,\vec{x}_1)
\Rightarrow \ldots \Rightarrow \phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ of
rewritings of the initial formula by substituting boolean states with
time-stamped transition rules. The word is accepted if the final
formula $\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ holds, when all
time-stamped variables $\vec{x}_1,\ldots,\vec{x}_n$ are substituted by
their values in $\nu_1,\ldots,\nu_n$, all non-final states replaced by
false and all final states by true.

The boolean operations of union, intersection and complement can be
implemented in linear time in this model, thus matching the complexity
of performing these operations in the finite-alphabet case. The price
to be paid is that emptiness becomes undecidable, for which reason we
provide two efficient semi-algorithms for emptiness, based on lazy
predicate abstraction \cite{blast} and the \impact~method
\cite{mcmillan06}. These algorithms are proven to terminate and return
a word from the language, if the language is not empty, but
termination is not guaranteed when the language is empty. Finally, we
prove termination in the empty case, when the transition
formulae\begin{inparaenum}[(i)]
\item contain only constants of the data sort and boolean functions of
  any arity, or they
%
\item constrain only the current but not the past values on the input,
  as in the case of symbolic automata \cite{symbTransd:POPL12,DAntoniKW16}.
\end{inparaenum}

We have implemented the boolean operations and emptiness checking
semi-algorithms in a prototype tool and carried out experiments with
examples taken from array logics \cite{cav09}, timed automata
\cite{henzinger:RealTimeSystems}, communication protocols \cite{abp}
and hardware verification \cite{smrcka}.

\paragraph{Related Work}
Data languages and automata have been defined previously, in a
classical nondeterministic setting. For instance, Kaminski and Francez
\cite{Kaminski94} consider languages, over an infinite alphabet of
data, recognized by automata with a finite number of registers, that
store the input data and compare it using equality. Just as the timed
languages recognized by timed automata \cite{AlurDill94}, these
languages, called quasi-regular, are not closed under complement, but
their emptiness is decidable. The impossibility of complementation
here is caused by the use of hidden variables, which we do not
allow. Even though closed under complement, emptiness is undecidable
in our case, mainly because counting (incrementing and comparing to a
constant) data values is allowed, in many data theories.

Other approaches to defining notions of data automata
\cite{Bojanczyk:2011:TLD,BouyerPetitTherien03} are different from ours
mainly because these automata are controlled by equivalence relations
of finite index, whereas in our case, the transitions are defined by
unrestricted formulae in the first-order theory of the data domain. As
in \cite{Kaminski94,AlurDill94}, the emptiness problem is decidable
for these classes.

Another related model is that of predicate automata
\cite{Farzan15,Farzan16}, which recognize languages over integer data
by labeling the words with conjunctions of uninterpreted
predicates. We intend to explore further the connection with our model
of alternating data automata, in order to apply our method to the
verification of parallel programs.

The model presented in this paper stems from the language inclusion
problem considered in \cite{Tacas16}. There we provide a
semi-algorithm for inclusion of data languages, based on an
exponential determinization procedure and an abstraction refinement
loop using lazy predicate abstraction \cite{blast}. In this work we
consider the full model of alternation and rely entirely on the
ability of SMT solvers to produce interpolants in the combined theory
of booleans and data. Since determinisation is not needed and
complementation is possible in linear time, the bulk of the work is
shipped to the decision procedure.

The emptiness check is implemented by adapting similar semi-algorithms
for nondeterministic infinite-state programs to the alternating model
of computation. In particular, we considered the state-of-the-art
\impact~ procedure \cite{mcmillan06} that is shown to outperform lazy
predicate abstraction in the nondeterministic case, and generalized it
to cope with alternation. More recent approaches for interpolant-based
abstraction refinement \cite{McMillan14,Hoder12} target Horn systems,
used to encode recursive and concurrent programs
\cite{Grebenshchikov12}. However, the emptiness of alternating word
automata cannot be directly encoded using Horn clauses, because all
the branches of the computation synchronize on the same input, which
does not seem to be easily encoded using a finite number of local
(equality) constraints. We believe that the lazy annotation techniques
for Horn clauses are suited for branching computations, which we
intend to consider in a future tree automata setting.
